Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}

### 2018-06-27 paper.tex
add a bunch of todos (mainly finding refs)

%% Title information
\title[Assessing primitives performance on multi-stage execution]{Assessing primitives performance\\ on multi-stage execution}

@@

On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation. Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method. We distinguish two kinds of primitives:
\begin{itemize}
	\item \emph{Essential primitives} cannot be performed in any other way. A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together. Many arithmetic and comparison operations between numbers are primitives. Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard.
	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}. They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
\end{itemize}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve.
	\item There is a big performance gap between the baseline performance and the peak performance.
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
\end{itemize}

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.

@@

The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

The loop overhead is measured and removed from the measurements (We made sure the JIT did not remove the dead loop to measure it). The two compared strings are equal as this situation corresponds to the worst scenario for most implementations. %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} shows the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.

@@

This difference is mainly due to the optimizations performed or not by both the compilers for the \legSlang version. As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar. On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in Figure \ref{fig:Asm}).

@@

\section{Discussion and related work}

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitives is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain.

>>>>

%% Title information
\title[Assessing primitives performance on multi-stage execution]{Assessing primitives performance\\ on multi-stage execution \todo{modify maybe}}

@@

On top of its virtual machine, the client language can use a set of \emph{primitives}\footnote{We use in the paper the Smalltalk terminology (primitive) as this is the programming language used for our evaluation. Some other programming languages, such as Javascript, prefer to use the term built-in instead of primitive.}, which are performed directly by the interpreter rather than by evaluating expressions in a method. We distinguish two kinds of primitives: \todo{motivate the distinction. See the algorithmic primitives paper. Is there a reference stating the different types of primitives ? }
\begin{itemize}
	\item \emph{Essential primitives} cannot be performed in any other way. A high-level object-oriented language without primitives can move values from one variable to another, but cannot add two integers together. Many arithmetic and comparison operations between numbers are primitives. Some primitives allow one to communicate with I/O devices such as the disk, the display and the keyboard.
	\item \emph{Optional primitives} exist only to make the system run faster \cite{blueBook}. They can be implemented in the client language directly, making its implementation as a primitive optional, but they are often implemented directly in the VM to improve performance.
\end{itemize}

@@

For this reason, some virtual machine implementors attempted to remove most or all optional primitives. Their goal was to use a JIT performing speculative optimizations to be able to implement the optional primitives in the client language without performance loss. Ideally they wanted to get   the performance of the same primitives written in the VM. Self \cite{ursPHD} and Strongtalk \cite{Strongtalk} were the first to try this approach. Early versions of the Javascript engine V8 \cite{V8} also implemented most of the primitives, including for example Array operations in Javascript itself. However, without optional primitives implemented in the VM, several problems rise:
\begin{itemize}
	\item High performance relies on a JIT with speculative optimizations, which is hard to implement, maintain and evolve. \todo{a ref would be nice}
	\item There is a big performance gap between the baseline performance and the peak performance. \todo{a ref would be nice}
	\item Even mature JITs fail to optimize some narrow cases, where the performance is drastically slower \\~\cite{VMHotCold}.
\end{itemize}

@@

\paragraph{String representations.} Strings are represented on top of the Cog VM in two possible forms: ByteStrings and WideStrings. ByteStrings encode each character of the string in a single byte. The encoding usually follows the extended ASCII standard. \todo{we should not use this term} WideStrings encode each character in 32 bits. The encoding usually follows Unicode specifications. In this paper, we will discuss only the implementation of the string comparison primitive for ByteStrings. WideStrings use an implementation entirely implemented in the client language, and so far, it has not been reported as a performance bottleneck for any program of any users. To sort arrays of strings, strings can be compared. They can be compared in the default order (ASCII or Unicode) or using different orders, for example, the case insensitive order.

@@

\todo{mention baseline here}
The \legSlangPlug, \legSlang and \legRTL implementations of this string comparison primitive have been implemented in the Cog VM. In this section, we compare execution speed of these primitives against the other existing implementations. We measure their respective execution times for different string sizes.% \sd{mention microbenchmarks here.}\sd{why there is only 4.3 and not something else? Because this is strange from a shape structure.}

@@

The loop overhead is measured and removed from the measurements (We made sure the JIT did not remove the dead loop to measure it). The two compared strings are equal as this situation corresponds to the worst scenario for most implementations.
\todo{explain how the relative speed up is computed} %There is one less branch in the two plugin implementations in this case.

\subsection{Results} 
We performed the benchmarks upon five different string sizes: 0, 1, 5, 100 and 1000 characters. In the case of small strings, most of the time is spent activating the primitive and in the code before the main comparison loop, the primitive is performed with a little overall number of native instructions. In the case of large strings, the time spent executing the primitive is dominated by the main comparison loop. Figure~\ref{fig:bench} and Table \ref{tab:detail_bench} show the results in terms of relative speed-up towards the baseline implementation, respectively for \GCCLinux and \LLVMMac.

@@

This difference is mainly due to the optimizations performed or not by the GCC and Clang compilers for the \legSlang version. As depicted in Figure~\ref{fig:Asm}, \GCCLinux generates instructions that are very close to the ones generated by the JIT, hence the execution time is similar. On the contrary, \LLVMMac generates more instructions: it fails at keeping all the operands live across the comparison loops, leading to two extra memory reads per iteration (bolded in Figure \ref{fig:Asm}).

@@

\section{Discussion and related work}
\todo{primitives in other system. (generalize more after Rpython. }

@@

The philosophy of Self and Strongtalk, both high performance VMs, were to improve the performance of the client language with a mature optimizing JIT featuring advanced optimizations such as speculative optimizations \cite{Strongtalk, ursPHD}. In this context, the need of primitives is reduced, since the performance gain from using a primitive over the client language is low or inexistant. More recently, Felgentreff \emph{et al.} \cite{NoPrimTracing} analysed the speed difference between primitives implemented in C, RPython, and Smalltalk. They showed that with the RPython toolchain framework, the speed difference between all three implementations is not that big, hence they could choose to implement some optional primitives in the client language over C or RPython without performance drop to reduce the number of primitives to maintain. \todo{emphasize the difference more}
